---
title: "Intro to MAF"
author: Nicole Black
date: 07/31/22
---

###### (1) Set Up

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/Users/joshuagabbay/Desktop/QBIO/qbio_490_joshuagabbay/analysis_data")) 
```

```{r}
# install and load maftools and TCGAbiolinks

if(!require("BiocManager", quietly = TRUE)){
  install.packages("BiocManager")
}
BiocManager::install("maftools")
library(maftools)

if(!require(TCGAbiolinks)){
  install.packages("TCGAbiolinks")
}

library(TCGAbiolinks)
```

Before we get into MAF, we need to read in the clinical data (created in week 4) so it can be merged with our MAF data later on.

```{r Exercise 1.1}
clinical <- read.csv("/Users/joshuagabbay/Desktop/QBIO/qbio_490_joshuagabbay/analysis_data/brca_clinical_data.csv")
```

We also need to change the bcr_patient_barcode column name that way the MAF package can read our clinical file.

```{r Exercise 1.2}
colnames(clinical)[ colnames(clinical) == "bcr_patient_barcode" ] <- "Tumor_Sample_Barcode"
```

After changing the column name, re-save the clinical data frame to your computer.

```{r}
# change the file path! make sure it's in your week4 folder
write.csv(clinical, "/Users/joshuagabbay/Desktop/QBIO/qbio_490_joshuagabbay/analysis_data/brca_clinical_data.csv", row.names = FALSE)
```

###### (2) MAF Package

Now, we can query in the MAF files using GDCquery()/GDCdownload()/GDCprepare() and read.maf(). 

```{r Exercise 2.1}
maf_query <- GDCquery(
    project = "TCGA-BRCA", 
    data.category = "Simple Nucleotide Variation", 
    access = "open", # we only have access to somatic mutations which are open access
    data.type = "Masked Somatic Mutation", 
    workflow.type = "Aliquot Ensemble Somatic Variant Merging and Masking"
)
GDCdownload(maf_query)
maf <- GDCprepare(maf_query) # as long as it runs, ignore any errors

maf_object <- read.maf(maf = maf, 
                       clinicalData = clinical,
                       isTCGA = TRUE)
```

###### (3) Exploring the MAF Object

Call str() on maf_object.

```{r Exercise 3.1}
# write code here
str(maf_object)
```

Notice the @ and $ signs. What do each of them represent?

  @ represents: the data frame you are in
  $ represents: the column you are in
  
List all of the data frames within the maf_object by typing maf_object@ then either waiting for a popup, or if no popup appears, by hitting tab.

```{r Exercise 3.2}
# write code here
maf_object@
```

Call head() and str() on the @data df and the @clinical.data df. These will be the main data frames we use in our MAF analysis.

```{r Exercise 3.3}
# write code here
head(maf_object@data)
str(maf_object@data)
head(maf_object@clinical.data)
str(maf_object@clinical.data)
```

View the first 110 items in Tumor_Sample_Barcode from @data and @clinical.data.

```{r Exercise 3.4}
# write code here
maf_object@data$Tumor_Sample_Barcode[1:110]
maf_object@clinical.data$Tumor_Sample_Barcode[1:110]
```

What do you notice about these two lists?

The first one is basically the same entry for the first 110 entries, while the second list is almost completely unique entries each time.

View the Tumor_Sample_Barcode column and the HGVSp_Short column for all TP53 mutations (i.e. Hugo_Symbol == "TP53"). Hint: boolean masking!

```{r Exercise 3.5}
# write code here
mask <- maf_object@data$Hugo_Symbol == "TP53"
maf_object@data$Tumor_Sample_Barcode[mask]
maf_object@data$HGVSp_Short[mask]
```

Call head() on the Amino_acids column in @data.

```{r Exercise 3.6}
# write code here
head(maf_object@data$Amino_acids)
```

What mutation could cause a S/R change?

G/U -> C/A in third spot

What about an A/T change?

GCU -> ACU

Call head() on the HGVSp_Short column in @data.

```{r Exercise 3.7}
# write code here
head(maf_object@data$HGVSp_Short)
```

What does p.P728Afs*11 mean? (Do some google-ing!)


```{r Exercise 3.8}
# use this code block for the following questions:
nrow(maf_object@data)
ncol(maf_object@data)
typeof(maf_object@data$Tumor_Sample_Barcode)
```
  
  How many rows/columns are there in @data?
  
  67254, 141
  
  What type of data is stored in the rows? What about the columns?
  
  list, integer
  
  How many levels are there in Tumor_Sample_Barcode (from @data)?
  Are there duplicate values in Tumor_Sample_Barcode?
  
  How many unique values are there in Amino_acids?
  Are there duplicate values in Amino_acids?
  
  How many unique values are there in HGVSp_Short?
  Are there duplicate values in HGVSp_Short?
