---
title: "Intro to MAF - Part II"
author: Nicole Black
date: 9/27/22
---

###### (1) Set Up

Set your path:

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/PATH/TO/analysis_data")) 
```

Load in necessary packages:

```{r}
# load maftools
```

Read in the clinical data.

```{r Exercise 1.1}
clinical <- read.csv("/PATH/TO/brca_clinical_data.csv")
```

Read in your maf object.

```{r Exercise 1.2}
maf_object <- read.maf(maf = FILL IN HERE, 
                       clinicalData = FILL IN HERE,
                       isTCGA = TRUE)
```

###### (2) Fisher's Exact Test - Demo

Rewrite the vital_status and breast_carcinoma_estrogen_receptor_status columns of your clinical df as factors.

```{r}
clinical$vital_status <- FILL IN HERE

clinical$breast_carcinoma_estrogen_receptor_status <- FILL IN HERE
```

Create a contingency table using table().

```{r}
contig <- table(FILL IN HERE)
```

Visualize our contingency table using a mosaic plot.

```{r}
mosaicplot(FILL IN HERE)
```

Run our Fisher's exact test.

```{r}
fisher_test <- fisher.test(FILL IN HERE)
```

View the results of our test.

```{r}
fisher_test

fisher_test$p.value
```

What does this p-value tell us?

Does this make sense given the biology of cancer?

###### (3) Mutational Co-Occurence Example

Subset your maf_object to create geneA_maf and geneB_maf.

```{r}
## half class do: TP53 + PIK3CA
## other half do: MUC16 + RYR2

geneA_maf <- subsetMaf(maf = FILL IN HERE,
                       genes = FILL IN HERE) ## fill in with your gene name as a string

geneB_maf <- subsetMaf(maf = FILL IN HERE,
                       genes = FILL IN HERE)
```

Now let's start to build our contingency table.

Access the barcodes of the patients with mutations in genes A and the barcodes of the patinets with mutations in gene B. HINT: Boolean indexing! How do we access patient barcodes within the MAF object without duplicate patients? @data or @clinical.data?

```{r}
mut_pats_geneA = FILL IN HERE
mut_pats_geneB = FILL IN HERE
```

Get the lengths of these two vectors.

```{r}
num_pats_geneA = length(FILL IN HERE)
num_pats_geneB = FILL IN HERE
```

Use intersect() to get the patients with mutations in both gene A AND gene B. Then find the length of that vector. What does this tell us in our contingency table?

```{r}
mut_pats_geneAB = intersect(FILL IN HERE)
num_AB = length(FILL IN HERE)
```

Now, determine the number of patients with one mutation and not the other. HINT: Do some subtraction with your above vectors.

```{r}
num_A_only = FILL IN HERE
num_B_only = FILL IN HERE
```

Finally, find the number of patients without either mutation. HINT: We know the total number of patients (based on our original maf_object), and we just calculated the number of patients with mutations (A only + B only + AB).

```{r}
num_neither = FILL IN HERE
```

Create your contingency table:

```{r}
contig <- matrix(c(num_AB, 
                         num_B_only,
                         num_A_only,
                         num_neither), 
                       nrow=2)

# view the contingency table
contig

# view the contingency table as a mosaic
mosaicplot(contig)
```

Run your Fisher exact test!

```{r}
fisher_test <- fisher.test(FILL IN HERE)
fisher_test
```

Are your mutations co-occurent or mutually exclusive?

###### (4) Maftools somaticInteractions()

Luckily, we don't have to do the above work by hand for each gene pair we want to look at because maftools has a function that does it for us!

Run somaticInteractions for your above genes by passing in a vector with your gene names into the genes argument.

```{r}
somaticInteractions(maf = FILL IN HERE,
                    genes = FILL IN HERE,
                    pvalue = c(0.05, 0.1)) ## these are the default p-value significance levels
```

Run the following line of code for the top 25 genes.

```{r}
somaticInteractions(maf = FILL IN HERE,
                    top = FILL IN HERE,
                    pvalue = c(0.05, 0.1))
```

Pick a gene pair that is either significantly co-occurant or significantly mutually exclusive and look up the genes! Does this make sense? Do you see any literature that expresses similar/oppositional views on these genes?

